---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "man/figures/README-",
    fig.align = "center",
    out.width = "100%",
    dpi = 75,
    collapse = TRUE,
    comment = "#>"
)
```

# interface

<!-- badges: start -->
[![R-CMD-check](https://github.com/dereckmezquita/interface/workflows/R-CMD-check/badge.svg)](https://github.com/dereckmezquita/interface/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/interface)](https://CRAN.R-project.org/package=interface)
[![GitHub version](https://img.shields.io/github/r-package/v/dereckmezquita/interface?label=GitHub)](https://github.com/dereckmezquita/interface)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Downloads](https://cranlogs.r-pkg.org/badges/interface)](https://cran.r-project.org/package=interface)
<!-- badges: end -->

The `interface` package provides a system for defining and implementing interfaces in R, with runtime type checking, bringing some of the benefits of statically-typed languages to R with zero dependencies.

`interface` provides:

1. **Interfaces**: Define and implement interfaces with type checking. Interfaces can be extended and nested.
1. **Typed Functions**: Define functions with strict type constraints.
1. **Typed Frames**: Choose between a `data.frame` or `data.table` with column type constraints and row validation.
1. **Enums**: Define and use enumerated types for stricter type safety.

## Installation

To install the package, use the following command:

```{r eval = FALSE}
# Install the package from the source
remotes::install_github("dereckmezquita/interface")
```

## Getting started

Import the package functions.

```{r}
box::use(interface[ interface, type.frame, fun, enum ])
```

Define an interface and implement it:

```{r}
# Define an interface
Person <- interface(
    name = character,
    age = numeric,
    email = character
)

# Implement the interface
john <- Person(
    name = "John Doe",
    age = 30,
    email = "john@example.com"
)

print(john)

# interfaces are lists
print(john$name)

# Modify the object
john$age <- 10
print(john$age)

# Invalid assignment (throws error)
try(john$age <- "thirty")
```

### Extending Interfaces and Nested Interfaces

Create nested and extended interfaces:

```{r}
# Define nested interfaces
Address <- interface(
    street = character,
    city = character,
    postal_code = character
)

Scholarship <- interface(
    amount = numeric,
    status = logical
)

# Extend interfaces
Student <- interface(
    extends = c(Address, Person),
    student_id = character,
    scores = data.table::data.table,
    scholarship = Scholarship
)

# Implement the extended interface
john_student <- Student(
    name = "John Doe",
    age = 30,
    email = "john@example.com",
    street = "123 Main St",
    city = "Small town",
    postal_code = "12345",
    student_id = "123456",
    scores = data.table::data.table(
        subject = c("Math", "Science"),
        score = c(95, 88)
    ),
    scholarship = Scholarship(
        amount = 5000,
        status = TRUE
    )
)

print(john_student)
```

### Custom Validation Functions

Interfaces can have custom validation functions:

```{r}
is_valid_email <- function(x) {
    grepl("[a-z|0-9]+\\@[a-z|0-9]+\\.[a-z|0-9]+", x)
}

UserProfile <- interface(
    username = character,
    email = is_valid_email,
    age = function(x) is.numeric(x) && x >= 18
)

# Implement with valid data
valid_user <- UserProfile(
    username = "john_doe",
    email = "john@example.com",
    age = 25
)

print(valid_user)

# Invalid implementation (throws error)
try(UserProfile(
    username = "jane_doe",
    email = "not_an_email",
    age = "30"
))
```

### Typed Functions

Define functions with strict type constraints:

```{r}
typed_fun <- fun(
    x = numeric,
    y = numeric,
    return = numeric,
    impl = function(x, y) {
        return(x + y)
    }
)

print(typed_fun(1, 2))  # [1] 3
try(typed_fun("a", 2))  # Invalid call
```

Functions with multiple possible return types:

```{r}
typed_fun2 <- fun(
    x = c(numeric, character),
    y = numeric,
    return = c(numeric, character),
    impl = function(x, y) {
        if (is.numeric(x)) {
            return(x + y)
        } else {
            return(paste(x, y))
        }
    }
)

print(typed_fun2(1, 2))  # [1] 3
print(typed_fun2("a", 2))  # [1] "a 2"
```

### Typed Data Frames and Data Tables

Create data frames with column type constraints and row validation:

```{r}
PersonFrame <- type.frame(
    frame = data.frame, 
    col_types = list(
        id = integer,
        name = character,
        age = numeric,
        is_student = logical
    )
)

# Create a data frame
persons <- PersonFrame(
    id = 1:3,
    name = c("Alice", "Bob", "Charlie"),
    age = c(25, 30, 35),
    is_student = c(TRUE, FALSE, TRUE)
)

print(persons)

# Invalid modification (throws error)
try(persons$id <- letters[1:3])
```

Additional options for data frame validation:

```{r}
PersonFrame <- type.frame(
    frame = data.frame,
    col_types = list(
        id = integer,
        name = character,
        age = numeric,
        is_student = logical,
        gender = enum("M", "F"),
        email = function(x) all(grepl("^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$", x))
    ),
    freeze_n_cols = FALSE,
    row_callback = function(row) {
        if (row$age >= 40) {
            return(sprintf("Age must be less than 40 (got %d)", row$age))
        }
        if (row$name == "Yanice") {
            return("Name cannot be 'Yanice'")
        }
        return(TRUE)
    },
    allow_na = FALSE,
    on_violation = "error"
)

df <- PersonFrame(
    id = 1:3,
    name = c("Alice", "Bob", "Charlie"),
    age = c(25, 35, 35),
    is_student = c(TRUE, FALSE, TRUE),
    gender = c("F", "M", "M"),
    email = c("alice@test.com", "bob_no_valid@test.com", "charlie@example.com")
)

print(df)
summary(df)

# Invalid row addition (throws error)
try(rbind(df, data.frame(
    id = 4,
    name = "David",
    age = 50,
    is_student = TRUE,
    email = "d@test.com"
)))
```

### Enums

Define enums for categorical variables; these are safe to use to protect a value from being modified to invalid options. The `enum` function creates a generator which is then used to create the enum object. This can be used standalone or as part of an interface.

```{r enum}
Colour <- enum("red", "green", "blue")

# Create an enum object
colour <- Colour("red")
print(colour)

colour$value <- "green"
print(colour)

# Invalid modification (throws error)
try(colour$value <- "yellow")

# Use in an interface
Car <- interface(
    make = enum("Toyota", "Ford", "Chevrolet"),
    model = character,
    colour = Colour
)

# Implement the interface
car1 <- Car(
    make = "Toyota",
    model = "Corolla",
    colour = "red"
)

print(car1)

# Invalid implementation (throws error)
try(Car(
    make = "Honda",
    model = "Civic",
    colour = "yellow"
))

# Invalid modification (throws error)
try(car1$colour$value <- "yellow")
try(car1$make$value <- "Honda")
```

## Conclusion

The `interface` package provides powerful tools for ensuring type safety and validation in R. By defining interfaces, typed functions, and typed data frames, you can create robust and reliable data structures and functions with strict type constraints. For more details, refer to the package documentation.

## Citation

If you use this package in your research or work, please cite it as:

Mezquita, D. (2024). interface: A Runtime Type System for R. R package version 0.1.0. https://github.com/dereckmezquita/interface
