---
title: "Introduction to interface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

The `interface` package provides a system for defining and implementing interfaces in R, with runtime type checking. This approach brings some of the benefits of statically-typed languages to R, allowing for more structured and safer code.

## Why Use Interfaces?

Interfaces in R can be beneficial for several reasons:

1. **Code Structure**: They provide a clear contract for what properties and methods an object should have.
1. **Type Safety**: They allow for runtime type checking, catching errors early.
1. **Documentation**: They serve as self-documenting code, clearly stating the expected structure of objects.
1. **Flexibility**: They allow for implementation of multiple interfaces, promoting code reuse.

`interface` package provides the following:

1. **Interfaces**: Define and implement interfaces with type checking. Interfaces can be extended and nested.
1. **Typed Functions**: Define functions with strict type constraints.
1. **Typed Data Frames**: Create data frames with column type constraints and row validation. Data frames can have custom validation functions and row callbacks.

## Installation

To install the package, use the following command:

```{r eval = FALSE}
# Install the package from the source
remotes::install_github("dereckmezquita/interface")
```

## Usage

Import the package functions.

```{r setup}
box::use(interface[ interface, type.frame, fun ])
```

### Interfaces

To define an interface, use the `interface()` function:

```{r}
# Define an interface
Person <- interface(
    name = character,
    age = numeric,
    email = character
)

# Implement the interface
john <- Person(
    name = "John Doe",
    age = 30,
    email = "john@example.com"
)

print(john)

# interfaces are lists
print(john$name)

# Valid assignment
john$age <- 10

print(john$age)

# Invalid assignment (throws error)
try(john$age <- "thirty")
```

### Nested Interfaces and Extending Interfaces

You can create nested interfaces and extend existing interfaces:

```{r}
# Define an Address interface
Address <- interface(
    street = character,
    city = character,
    postal_code = character
)

# Define a Scholarship interface
Scholarship <- interface(
    amount = numeric,
    status = logical
)

# Extend the Person and Address interfaces
Student <- interface(
    extends = c(Address, Person), # will inherit properties from Address and Person
    student_id = character,
    scores = data.table::data.table,
    scholarship = Scholarship # nested interface
)

# Implement the extended interface
john_student <- Student(
    name = "John Doe",
    age = 30,
    email = "john@example.com",
    street = "123 Main St",
    city = "Small town",
    postal_code = "12345",
    student_id = "123456",
    scores = data.table::data.table(
        subject = c("Math", "Science"),
        score = c(95, 88)
    ),
    scholarship = Scholarship(
        amount = 5000,
        status = TRUE
    )
)

print(john_student)
```

### Custom Validation Functions

Interfaces can have custom validation functions:

```{r}
is_valid_email <- function(x) {
    grepl("[a-z|0-9]+\\@[a-z|0-9]+\\.[a-z|0-9]+", x)
}

UserProfile <- interface(
    username = character,
    email = is_valid_email,
    age = function(x) is.numeric(x) && x >= 18
)

# Implement with valid data
valid_user <- UserProfile(
    username = "john_doe",
    email = "john@example.com",
    age = 25
)

print(valid_user)

# Invalid implementation (throws error)
try(UserProfile(
    username = "jane_doe",
    email = "not_an_email",
    age = "30"
))
```

### Typed Functions

Define functions with strict type constraints:

```{r}
typed_fun <- fun(
    x = numeric,
    y = numeric,
    return = numeric,
    impl = function(x, y) {
        return(x + y)
    }
)

# Valid call
print(typed_fun(1, 2))  # [1] 3

# Invalid call (throws error)
try(typed_fun("a", 2))
```

#### Functions with Multiple Possible Return Types

```{r}
typed_fun2 <- fun(
    x = c(numeric, character),
    y = numeric,
    return = c(numeric, character),
    impl = function(x, y) {
        if (is.numeric(x)) {
            return(x + y)
        } else {
            return(paste(x, y))
        }
    }
)

print(typed_fun2(1, 2))  # [1] 3
print(typed_fun2("a", 2))  # [1] "a 2"
```

### Typed `data.frame`/`data.table`s

Create data frames with column type constraints and row validation:

```{r}
PersonFrame <- type.frame(
    frame = data.frame, 
    col_types = list(
        id = integer,
        name = character,
        age = numeric,
        is_student = logical
    )
)

# Create a data frame
persons <- PersonFrame(
    id = 1:3,
    name = c("Alice", "Bob", "Charlie"),
    age = c(25, 30, 35),
    is_student = c(TRUE, FALSE, TRUE)
)

print(persons)

# Invalid modification (throws error)
try(persons$id <- letters[1:3])
```

#### Additional Options

```{r}
PersonFrame <- type.frame(
    frame = data.frame,
    col_types = list(
        id = integer,
        name = character,
        age = numeric,
        is_student = logical,
        email = function(x) all(grepl("^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$", x))
    ),
    freeze_n_cols = FALSE,
    row_callback = function(row) {
        if (row$age >= 40) {
            return(sprintf("Age must be less than 40 (got %d)", row$age))
        }
        if (row$name == "Yanice") {
            return("Name cannot be 'Yanice'")
        }
        return(TRUE)
    },
    allow_na = FALSE,
    on_violation = "error"
)

df <- PersonFrame(
    id = 1:3,
    name = c("Alice", "Bob", "Charlie"),
    age = c(25, 35, 35),
    is_student = c(TRUE, FALSE, TRUE),
    email = c("alice@test.com", "bob_no_valid@test.com", "charlie@example.com")
)

print(df)
summary(df)

# Invalid row addition (throws error)
try(rbind(df, data.frame(
    id = 4,
    name = "David",
    age = 500,
    is_student = TRUE,
    email = "d@test.com"
)))
```

## Conclusion

This package provides powerful tools for ensuring type safety and validation in R. By defining interfaces, typed functions, typed data frames, and enums, you can create robust and reliable data structures and functions with strict type constraints.

This vignette demonstrates the basic usage and capabilities of the package. For more details, refer to the package documentation and examples provided in the source code.
